---
title: "Assignment 4"
author: "Fabrizio Fiorini"
date: "3/21/2021"
output: html_document
---


## LOADING AND EXPLORATION

In this assignment I am going to perform a cluster analysis applying a type of non-hierarchical algorithm, k-means clustering algorithm.
The goal is to segment the data into a set of homogeneous clusters that will allow us to draw valuable information.

As a first step, let us load the required packages and the original dataset. It contains information about 21 pharmaceutical companies traded in 3 different stock exchanges.

```{r results='hide'}
#loading the packages
library(caret)
library(factoextra)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(cowplot)

#exploring the dataset
pharma <- read.csv("Pharmaceuticals.csv")
head(pharma)
str(pharma)
summary(pharma)
dim(pharma)
colMeans(is.na(pharma))

#giving names to rows as "Name" variable, then deleting the column
row.names(pharma) <- pharma[,2]
pharma <- pharma[,-2]
```

For the first part of the assignment I will focus on a subset of the original dataset that consists only of numerical variables.

```{r}
#excluding "Symbol" and the last 3 non-numerical variables
pharma.short <- pharma[,-c(1,11:13)]
```

## NORMALIZATION AND CLUSTERING

Here I measure the distance between each observation. Since by default the Euclidean distance metric is applied and it is sensitive to scale, data needs to be normalized first.
From the chart we can see different intensity of color for different distances. As we would expect, the diagonal has a value equal to zero because it indicates the distance of an observation from itself.

```{r}
#normalizing data
norm.pharma.short <- scale(pharma.short)

#measuring and plotting distance
distance <- get_dist(norm.pharma.short)
fviz_dist(distance)
```

One of the most helpful tools that are used to choose the number of cluster for the k-means model when there are not any external considerations are the Elbow chart and the Silhouette Method. The former shows the decline in cluster heterogeneity when we add more clusters. The latter measures how similar an object is to its cluster compared to the other clusters.
From the charts plotted, we can see that the elbow in the line is when k=2, while the Silhouette Method gives us k=5 as a result.
I decided to run the k-means algorithm with 5 clusters.

```{r}
#plotting elbow chart and silhouette method
Wss <- fviz_nbclust(norm.pharma.short, kmeans, method = "wss")
Sil <- fviz_nbclust(norm.pharma.short, kmeans, method = "silhouette")
plot_grid(Wss, Sil)

#running k-means with k=5
set.seed(123)
km.pharma.best <- kmeans(norm.pharma.short, centers = 5, nstart = 50)
km.pharma.best$centers
km.pharma.best$size
km.pharma.best$withinss
fviz_cluster(km.pharma.best, data = norm.pharma.short)
```

From the result we can characterized the 5 clusters looking at the distance from the centers. For example, cluster n.4 has a high Market Capital, cluster n.2 has high Beta, cluster n.5 has a low Asset Turnover, and so on.
We can also retrieve the size of each cluster. Cluster n.1 is the largest, while cluster n.3 has only 2 firms.
The within-cluster sum of squared distances tell us about the dispertion of data: cluster n.1 (21.9) is less homogeneous than cluster n.3 (2.8).
By plotting the result of the algorithm we see the 5 group in which data has been clustered.


For the sake of a better interpretation of the cluster analysis, I decided to run again the model using only 3 clusters, since with only 2 clusters we could lost some of the characteristics of the data.
In this way, it is easier to identify and manage the cluster in the analysis. We now have 4 data points in cluster n.1, 11 in cluster n.2, and 6 in cluster n.3.


```{r}
#running k-means algorithm with k=3
set.seed(123)
km.pharma <- kmeans(norm.pharma.short, centers = 3, nstart = 50)
km.pharma$centers
km.pharma$size
km.pharma$withinss
fviz_cluster(km.pharma, data = norm.pharma.short)
```

```{r echo=FALSE}
#plotting data points grouped in clusters
centroids <- data.frame(km.pharma$centers) %>% 
  rowid_to_column() %>% 
  gather('Columns', 'Centers', -1)
ggplot(centroids, aes(x = Columns, y = Centers, color = as.factor(rowid))) + 
  geom_line(aes(group = as.factor(rowid))) + geom_point()
```

From the second chart we see that cluster n.1 is formed by companies with a low Net Profit Margin and a high Price/Earnings ratio, while firms in cluster n.2 have low Asset Turnover and Return on Asset (ROA) but high Leverage and Estimated Revenue Growth. Cluster n.3 does not stand out in any of the metrics we have analyzed.

## CATEGORICAL VARIABLES

Let us now consider the last 3 categorical variables: Median Recommendation, Location and Stock Exchange.
In order to look for possible trends within data, I decided to leverage bar charts to graphically visualize the distribution of the firms grouped by clusters.
From the plotting we can clearly see that the majority of the companies in cluster n.3 are located within the US and they share a spread recommendation of holding their shares. All of them are traded on the NYSE. In cluster n.2 we have a preference for "Moderate Buy" shares and the only 2 companies whose stocks are traded in other markets or indexes (AMEX and NASDAQ) are contained here. In cluster n.1 we can notice that the 4 firms are located in 4 different countries while their stocks are traded on the NYSE.

```{r}
#partitioning data set
pharma.cat <-  pharma %>% select(c(11,12,13)) %>% 
    mutate(Cluster = km.pharma$cluster)
```

```{r echo=FALSE}
#plotting clusters
Rec <- ggplot(pharma.cat, mapping = aes(factor(Cluster), fill=Median_Recommendation)) +
  geom_bar(position = 'dodge') +
  labs(x='Clusters', y='Frequence')

Loc <- ggplot(pharma.cat, mapping = aes(factor(Cluster), fill=Location)) +
  geom_bar(position = 'dodge') + 
  labs(x='Clusters', y='Frequence')

Exc <- ggplot(pharma.cat, mapping = aes(factor(Cluster), fill=Exchange)) +
  geom_bar(position = 'dodge') + 
  labs(x='Clusters', y='Frequence')

plot_grid(Rec, Loc, Exc)
```


## FINAL THOUGHTS - Characterize the 3 clusters

Finally, we are able now to gather all the insights we could retrieve from the dataset and identify 3 distinct groups from the basket of 21 pharmaceutical companies.
Cluster n.1 can be labeled as "overvalued international firms" due to the follow characteristics: international location, traded on the NYSE, low Net Profit Margin and a high Price/Earnings ratio. These companies have their businesses spread across multiple continents and at the same time raise capital from the world largest stocks exchange (NYSE). They share a high valuation on the financial market not justified by the current level of profit. They need to meet investors' expectations by investing and increasing profits if they do not want their stock price to decrease.
Cluster n.2 can be labeled as "growing and leveraged firms" due to the follow characteristics: "Moderate buy" recommendation, low Asset Turnover and ROA, high Leverage and Estimated Revenue Growth. They seems to be higly appreciated from investors that are willing to wait for future growth despite the current low profits and high debt.
Cluster n.3 can be labeled as "mature US firms" due to the follow characteristics: US-based, traded on the NYSE, "Hold" recommendations.